---
title: "1. Generative trees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Generative trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Every simulation in Smith (2026) is generated along one of 250 "true" trees,
so that reconstruction accuracy can be judged by comparison back to a known
answer. This vignette explains where those trees came from, and how to
regenerate (or extend) the sample.

## The bundled trees

Regenerating the tree posterior takes hours (see below), so the actual
250-tree sample used throughout the package is bundled directly:

```{r}
library(PhyloPCA)
data(generative_trees)
generative_trees
```

```{r, fig.width = 6, fig.height = 6}
plot(generative_trees[[1]], cex = 0.6, no.margin = TRUE)
```

Trees are indexed stably: `generative_trees[[k]]` always refers to the same
tree, however many trees are sampled in total (see "How the sample is
built", below) -- so a tree index used elsewhere in this package (e.g.
`scale$tree`, `shape_combine_nj$t`) unambiguously identifies one of these
trees.

## Where they come from

The trees are a posterior sample from a Bayesian re-inference of
Mongle et al. (2023)'s discrete character matrix (15 hominins, 6 anthropoid
outgroups, 107 characters), under the Mk model (Lewis 2001), matching
Raskin et al. (2026)'s own inference model exactly except for the
ascertainment-bias correction: `coding = "variable"`, the correct choice
for a matrix with no invariant characters. A small excerpt of the matrix is
bundled for illustration:

```{r}
mongleFile <- system.file("extdata", "mongle_2023.nex", package = "PhyloPCA")
MongleStateProportions(mongleFile)
```

This tabulates, for each observed number of states `k`, how many of the
107 characters attain it -- the distribution `SimulateDiscreteMS()` draws
from when simulating new discrete characters (see
`vignette("simulating-data")`).

### Re-running the re-inference (not required)

`RunRevBayesReinference()` runs this model via RevBayes. It takes hours even
for 21 taxa, so this chunk does not run automatically:

```{r, eval = FALSE}
result <- RunRevBayesReinference(
  nexusFile = mongleFile,
  outDir = tempfile("reinfer")
)
result$runFiles # the two independent .trees traces
```

### From trace to tree sample

Given two converged tree traces, `SampleStableTrees()` draws a fixed,
append-only sample: trees from both runs are interleaved (so both chains
contribute from the very first draw) and then strided, so that `tree_k` is
a fixed function of `k` -- raising the sample size later only appends new
trees, never redraws existing ones. Convergence is checked first via
`TreeESS()`, the topological effective-sample-size diagnostic (the lower
of the Frechet-correlation and median-pseudo ESS statistics; Fabreti &
Hohna 2022; Magee et al. 2024) used throughout Smith (2026):

```{r, eval = FALSE}
runFiles <- result$runFiles
trees250 <- SampleStableTrees(runFiles, nTrees = 250, stride = 11)
```

`stride = 11` is large enough that consecutive draws from the interleaved
pool are effectively independent; `SampleStableTrees()` errors informatively
if a larger sample is requested than the trace supports, rather than
silently reusing correlated draws.
