---
title: "3. PCA, alignment and distances"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. PCA, alignment and distances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(PhyloPCA)
data(generative_trees)
truth <- generative_trees[[1]]
```

## Continuous data: principal components and distance

For continuous characters, `prcomp()` (base R) already does everything
needed: Smith (2026) centres each character on its mean, without rescaling
to unit variance, then computes Euclidean distance on (i) the first two
principal components (Raskin et al. 2026's pipeline); (ii) all principal
components (equivalent to distance on the original scaled characters); and
(iii) the *additive* (squared-Euclidean) distance, appropriate for
neighbour-joining because the expected **squared** distance between taxa
under Brownian motion is proportional to tree (patristic) distance:

```{r}
set.seed(1)
M <- SimulateContinuous(truth, nChar = 100)
scores <- prcomp(M)$x

dPipeline <- dist(scores[, 1:2]) # Raskin et al. (2026)'s pipeline
dAllPC <- dist(scores) # all dimensions retained
dAdditive <- SquaredEuclideanDistance(scores) # the additive fix
```

Discrete characters get the analogous additive fix via
`CorrectedDistanceMS()` (a per-character-state-size Jukes-Cantor
correction; reduces to the familiar binary correction when every character
has two states), compared to plain `HammingDistance()`:

```{r}
mongleFile <- system.file("extdata", "mongle_2023.nex", package = "PhyloPCA")
stateProportions <- MongleStateProportions(mongleFile)
X <- SimulateDiscreteMS(truth, nChar = 100, stateProportions = stateProportions)
dHamming <- HammingDistance(X)
dCorrected <- CorrectedDistanceMS(X)
```

## Shape data: superimposition

Landmark configurations must be superimposed (aligned) before their
coordinates are comparable between taxa. `PhyloPCA` provides two
alignments, both dependency-free, pure-R implementations:

- `ProcrustesAlign()`: ordinary least-squares generalized Procrustes
  analysis (GPA) -- centre, scale to unit centroid size, iteratively rotate
  onto an evolving consensus. Equivalent to `geomorph::gpagen()` for
  complete landmark data.
- `RftraAlign()`: resistant-fit (RFTRA) superimposition (Siegel & Benson
  1982; Slice 1996) -- a repeated-median scale and Tukey-bisquare
  reweighted rotation, so that a small number of grossly displaced
  landmarks (the "Pinocchio effect": Palci & Lee 2019) are down-weighted
  rather than smeared across the whole configuration.

```{r}
sampleFile <- system.file("extdata", "sample_shape_t00_d2_s0.4.tsv", package = "PhyloPCA")
configs <- ReadShapeConfigs(sampleFile, dim = 2)[truth$tip.label] # order to tree tips
aligned <- ProcrustesAlign(configs)
alignedRftra <- RftraAlign(configs)
```

Both alignments return a named list of landmarks x dim matrices, in the
same order supplied. To compute a distance for neighbour-joining, flatten
each aligned configuration to one row per taxon:

```{r}
flat <- t(sapply(aligned, function(z) c(z)))
dShape <- dist(flat)
```

Which alignment is better depends on the shape kernel: Smith (2026) finds
Procrustes and RFTRA barely distinguishable in practice (the LDDMM
deformations used are smooth and distributed, without the isolated
"Pinocchio" landmark outliers RFTRA is designed to resist), and both
noticeably better than no superimposition once the shape kernel is
sufficiently integrated (see `vignette("scoring-and-figures")` and the
`shape_sigma_nj_align`/`score_native`/`score_rftra` datasets, which compare
all three).

Continuing this worked example through to a scored neighbour-joining tree
is shown next, in `vignette("tree-inference")`.
