---
title: "5. Scoring reconstructions and reproducing the figures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Scoring reconstructions and reproducing the figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4.5)
```

```{r setup}
library(PhyloPCA)
```

## Scoring one reconstruction

`ScoreToTruth()` computes the three tree-distance metrics used in
Smith (2026): clustering information distance (CID; Smith 2020, the primary
metric -- normalized 0-1, ~0.84 expected between random 21-tip trees),
Robinson-Foulds distance (for comparability with Raskin et al. 2026), and
(where both trees are strictly binary) SPR distance. `ExactlyRecovered()`
reports the (very strict, for 21 tips) binary criterion of an exact
topology match:

```{r}
data(generative_trees)
truth <- generative_trees[[1]]
guess <- ape::rtree(21, tip.label = truth$tip.label) # a random tree, for illustration

ScoreToTruth(guess, truth, metrics = c("CID", "RF"))
ExactlyRecovered(guess, truth)
```

`OptimalityRecovery()` implements the alternative, parsimony-specific
recovery criterion used for the maximum-parsimony arms: whether the
generative tree is *tied for most parsimonious*, rather than whether a
search happened to find it -- so a heuristic search's failure to locate the
true topology among several equally-parsimonious trees is not counted
against the parsimony criterion itself.

## The cached results

Every number behind the manuscript's figures is bundled as package data.
`?PhyloPCA::scale` and its neighbours document exactly what each dataset
contains and which script produced it (see the "Data" index at
<https://phylo-pca.github.io/reference/>). At a glance:

```{r}
data(scale) # continuous + discrete, NJ + Bayesian, nC 25-250
data(scale_fig1_dists) # 5 method arms on continuous data (Figure 1)
head(scale_fig1_dists)
```

## Reproducing Figure 1: which reconstruction method?

Figure 1 of Smith (2026) compares five ways of reconstructing a tree from
*the same* continuous character matrices: the PCA -> neighbour-joining
pipeline (Raskin et al. 2026's method), neighbour-joining on all principal
components, neighbour-joining on the additive (squared-Euclidean) distance,
maximum parsimony, and Bayesian inference. A simplified, base-graphics
version of that comparison, drawn directly from the cached data:

```{r fig1}
arms <- c(pipeline = "PCA -> NJ", allpc = "NJ, all PCs",
          additive = "NJ, additive", mp = "Parsimony", bayes = "Bayesian")
col <- c("#E69F00", "#56B4E9", "#009E73", "#D55E00", "#000000")

meanBy <- function(arm) {
  d <- scale_fig1_dists[scale_fig1_dists$arm == arm, ]
  sapply(split(d$CID, d$nC), mean)
}
nC <- sort(unique(scale_fig1_dists$nC))
plot(NA, xlim = range(nC), ylim = c(0, 0.35), log = "x", las = 1,
     xlab = "Number of characters", ylab = "Mean CID to true tree")
for (i in seq_along(arms)) {
  lines(nC, meanBy(names(arms)[i]), col = col[i], lwd = 2, type = "b", pch = 16)
}
legend("topright", legend = arms, col = col, lwd = 2, pch = 16, bty = "n")
```

The pipeline arm (Raskin et al. 2026's own method) is the clear outlier;
correcting only the *distance* fed to neighbour-joining (the additive arm)
recovers essentially all of the gap to Bayesian inference, without changing
the reconstruction method at all.

## The full manuscript figures

The three figures actually published with Smith (2026) (`fig_methods.png`,
`fig_alignment.png`, `fig_data.png`) add production styling -- a
colourblind-safe palette, print-sized multi-panel layouts, off-scale
annotations -- that is beyond this package's scope. That styling is the
manuscript repository's concern, not this package's; what belongs here, and
is complete, is every number those figures plot: each is one of the columns
documented in the datasets above, produced by the functions in this
package.
