---
title: "2. Simulating character data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Simulating character data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(phyloPCA)
data(generative_trees)
truth <- generative_trees[[1]]
```

Smith (2026) simulates three kinds of character data along each generative
tree: continuous morphological characters, discrete morphological
characters, and geometric-morphometric ("shape") landmark data. All three
are compared on an equal footing at the end of the analysis (see
`vignette("scoring-and-figures")`).

## Continuous characters

Continuous characters are simulated under Brownian motion (root state 0,
rate `sigma^2 = 1`, following Raskin et al. 2026), independently for each
character:

```{r}
set.seed(1)
continuous <- SimulateContinuous(truth, nChar = 50)
dim(continuous)
continuous[1:4, 1:4]
```

## Discrete characters

Discrete characters are simulated under the Mk model (Lewis 2001), with no
rate variation among characters. Each character's number of possible states
is drawn from the empirical distribution observed in Mongle et al. (2023)'s
matrix (see `vignette("generative-trees")`), and characters that do not
vary across tips are rejected and re-drawn (`coding = "variable"`), matching
the ascertainment correction applied at inference time:

```{r}
mongleFile <- system.file("extdata", "mongle_2023.nex", package = "phyloPCA")
stateProportions <- MongleStateProportions(mongleFile)
set.seed(1)
discrete <- SimulateDiscreteMS(truth, nChar = 50, stateProportions = stateProportions)
dim(discrete)
discrete[1:4, 1:8]
```

Every column is variable across tips:

```{r}
all(apply(discrete, 2, function(col) length(unique(col)) > 1))
```

## Shape (geometric-morphometric) data

Shape data are simulated with the bespoke LDDMM (large deformation
diffeomorphic metric mapping) shape simulator of Raskin et al. (2026): a
circular (2D) or spherical (3D) template of evenly spaced landmarks evolves
along each branch under an Euler-Maruyama discretisation of a Matern-like
spatial kernel, rejecting any step that would cause landmarks to cross.

**This package does not vendor that simulator.** Raskin et al. (2026)'s
repository (<https://github.com/Levi-Raskin/PCAPhylogenetics>) carries no
license, so redistributing their C++ source here would not be a
reproducibility aid worth the ambiguity it creates. To regenerate shape data
from scratch: clone their repository, build `ShapeSimulationCode` (a
recompiled, byte-for-byte-faithful harness is described in the `phaseC/`
directory of the analysis repository at
<https://github.com/phylo-pca/phylo-pca.github.io>), and run it on the trees
in `generative_trees`.

Because most users will never need to run the external simulator, a small
sample of its output is bundled so that everything downstream of simulation
-- alignment, tree inference, scoring -- can be exercised and verified
end-to-end without it:

```{r}
sampleFile <- system.file("extdata", "sample_shape_t00_d2_s0.4.tsv", package = "phyloPCA")
configs <- ReadShapeConfigs(sampleFile, dim = 2)
length(configs) # one landmark configuration per taxon
dim(configs[[1]]) # landmarks x dimensions
```

This particular sample is 2D, 25-landmark data at a shape kernel
`sigma = 0.4`, for the tree `generative_trees[[1]]` (`t00` in the 0-based
indexing used by the simulation grid). See
`vignette("pca-and-alignment")` for what to do with it next.
