---
title: "4. Tree inference: neighbour-joining, parsimony, Bayesian"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Tree inference: neighbour-joining, parsimony, Bayesian}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
hasTNT <- tryCatch({ FindTNT(); TRUE }, error = function(e) FALSE)
hasRB <- tryCatch({ phyloPCA::FindRevBayes(); TRUE }, error = function(e) FALSE)
```

```{r setup}
library(phyloPCA)
library(ape)
data(generative_trees)
truth <- generative_trees[[1]]
set.seed(1)
M <- SimulateContinuous(truth, nChar = 100)
scores <- prcomp(M)$x
```

Smith (2026) reconstructs a tree from every simulated dataset by three
different methods, so that reconstruction accuracy can be attributed to the
*data* or to the *method* used to analyse it. This vignette covers all
three for continuous data; the same functions apply to discrete and (for
parsimony/Bayesian) shape data, with `type`/template arguments changed as
noted.

## Neighbour-joining

Neighbour-joining (`ape::nj()`) needs no wrapper -- this package's
contribution is providing the correct *distance* to feed it (see
`vignette("pca-and-alignment")`):

```{r}
treeNJ <- nj(SquaredEuclideanDistance(scores))
ScoreToTruth(treeNJ, truth, metrics = c("CID", "RF"))
```

## Maximum parsimony (TNT)

`RunTNTParsimony()` runs the unified search protocol validated across every
parsimony arm in Smith (2026): 50 random-addition-sequence + TBR starting
trees, a New Technology pass (sectorial search, ratchet, drift, tree
fusing -- confirms rather than improves on the RAS+TBR optimum here), then
TBR-swaps every optimal tree to completion and saves all most-parsimonious
trees. `type = "continuous"` treats the data as Wagner-optimized real-valued
characters (Goloboff et al. 2006); `"discrete"` as standard (Fitch)
unordered characters; `"landmark"` as linear-displacement-optimized
landmark configurations (Goloboff & Catalano's method, following Palci &
Lee 2019).

This needs a local TNT installation (`options(phyloPCA.tnt = ...)` or the
`TNT_PATH` environment variable; see `FindTNT()`) -- `r if (!hasTNT) "not available when this vignette was built, so this chunk is shown but not run" else "available, so this chunk actually ran"`:

```{r, eval = hasTNT}
result <- RunTNTParsimony(M, type = "continuous")
result$nTree # essentially always 1 for continuous (real-valued) data
ScoreMPTsByMaxEntropy(result$trees, truth)
```

Where a search finds more than one most-parsimonious tree (routine for
discrete and shape data), a strict consensus is over-conservative --
`ScoreMPTsByMaxEntropy()` instead takes a maximum-entropy subsample of up
to `k` trees (via `MaxMin::MaxEntropy()`, using clustering information
distance as the measure of redundancy between trees) and reports their mean
distance to the truth.

## Bayesian inference (RevBayes)

`RunRevBayesBM()` and `RunRevBayesMk()` run the two-run,
Metropolis-coupled Bayesian analyses used throughout Smith (2026): a
Brownian motion model (`dnPhyloBrownianREML`) for continuous/shape data, and
a per-state-size Mk model for discrete data. Both stop themselves once the
Gelman-Rubin PSRF falls below 1.01 and the parameter ESS exceeds 333.

This needs a local RevBayes installation (see `FindRevBayes()`) and takes
on the order of a minute even for this small example -- `r if (!hasRB) "not available when this vignette was built, so this chunk is shown but not run" else "available, so this chunk actually ran"`:

```{r, eval = hasRB}
nexusFile <- tempfile(fileext = ".nex")
WriteContinuousNexus(M, nexusFile)
outputPrefix <- tempfile("bm")
result <- RunRevBayesBM(nexusFile, outputPrefix, seed = 1)
result$mapTrees

mapTree <- ReadMAPTree(result$mapTrees[1])
ScoreToTruth(mapTree, truth, metrics = c("CID", "RF"))
```

`RunRevBayesMk()` works the same way, from a discrete-character Nexus file
(`WriteDiscreteNexus()`) instead.

## Putting a number on convergence

Alongside a point estimate (the MAP tree above), Smith (2026) reports a
Gelman-Rubin PSRF and a topological effective sample size for every
Bayesian analysis; see `TreeESS()` in `vignette("generative-trees")` for
the latter. Because RevBayes' own trace files (`<prefix>_run_1.trees`,
`<prefix>_run_2.trees`) are exactly the input `TreeESS()` expects, the same
diagnostic used to certify the generative-tree posterior applies unchanged
to every inference run in the pipeline.
